---
title: "R Laboratory – Exercise 1"
author: "Guglielmo Bordin"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        theme: readable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# American Airlines Employees
## Reading Data
We begin by importing the `tidyverse` and setting `ggplot`-related stuff.
```{r}
library(tidyverse)
font <- "Fira Sans Condensed"
theme_set(theme_bw(base_size = 20, base_family = font))
```
Now we can start importing data. Let’s first put each company’s data in a
separate tibble, and then merge them in a single one, adding a column with
the company name. Also, we’ll combine the `month` and `year` columns into 
a single date column (labeled as `month`).
```{r}
names <- c("month", "year", "full_time", "part_time", "total")
american <- read_table("data/american_airline_empl.txt",
                       skip = 1, col_names = names)
delta <- read_table("data/delta_airline_empl.txt",
                    skip = 1, col_names = names)
federal <- read_table("data/federal_express_empl.txt",
                      skip = 1, col_names = names)
united <- read_table("data/united_airline_empl.txt",
                      skip = 1, col_names = names)

# start with a named list to recover the names when binding rows
(employees <- list(american = american, delta = delta,
                   federal = federal, united = united) |>
    bind_rows(.id = "company") |>
    mutate(month = lubridate::make_date(year = year, month = month)) |>
    select(-year))
```
We’ll also keep track of properly-typeset company and job names, to use 
in the plot labels later.
```{r}
full_names <- list(american = "American Airlines",
                   delta = "Delta Airlines",
                   federal = "Federal Express",
                   united = "United Airlines")
job_type <- list(full_time = "Full-Time",
                 part_time = "Part-Time",
                 total = "Total")
```
