---
title: R Laboratory – Exercise 1
author: Guglielmo Bordin
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output:
    rmdformats::robobook:
        fig_width: 8
        fig_height: 5
        thumbnails: false
        lightbox: true
        gallery: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, dev = "svg")
```

# American Airlines Employees
## Reading Data
We begin by importing the necessary packages and setting `ggplot`-related
stuff.

```{r}
library(kableExtra)
library(tidyverse)

font <- "Roboto Condensed"
theme_set(theme_minimal(base_size = 14, base_family = font))
my_palette <- unname(palette.colors(4, "Okabe-Ito"))
```

Now we can start importing data. Let’s first put each company’s data in a
separate tibble, and then merge them in a single one, adding a column with
the company name. Also, we’ll combine the `month` and `year` columns into 
a single date column (labeled as `month`).

```{r}
names <- c("month", "year", "full_time", "part_time", "total")
american <- read_table("data/american_airline_empl.txt",
                       skip = 1, col_names = names)
delta <- read_table("data/delta_airline_empl.txt",
                    skip = 1, col_names = names)
federal <- read_table("data/federal_express_empl.txt",
                      skip = 1, col_names = names)
united <- read_table("data/united_airline_empl.txt",
                      skip = 1, col_names = names)

# start with a named list to recover the names when binding rows
employees <- list(american = american, delta = delta,
                   federal = federal, united = united) |>
    bind_rows(.id = "company") |>
    mutate(month = make_date(year = year, month = month)) |>
    select(-year)
```
```{r, echo = FALSE}
kbl(head(employees), caption = "`employees` tibble")
```

We’ll also keep track of properly-typeset version of the company and jobs
names, to use them later in the plots.

```{r}
full_names <- list(american = "American Airlines",
                   delta = "Delta Airlines",
                   federal = "Federal Express",
                   united = "United Airlines")
job_type <- list(full_time = "Full-time",
                 part_time = "Part-time",
                 total = "Total")
```

## Full-Time vs Part-Time Employees
Here we’ll generate a plot of the number of employees in time, for all 
four companies. We’ll keep the full-time and part-time workers’ numbers
separated, and also produce a plot with the total.

```{r}
lapply(names(job_type), function(idx) {
    employees |>
    ggplot(aes(x = month, y = .data[[idx]], colour = company)) +
        geom_line(lwd = 0.8) +
        scale_colour_manual(values = my_palette, labels = full_names) +
        labs(x = "Year", y = "Number of employees", colour = "Company",
             title = paste(job_type[[idx]], "Employees"))
})
```

To answer the question of when did each company reach the minimum and 
maximum number of employees, we can directly work on `employees` with the 
`dplyr` functions `group_by` and `summarize`.

```{r}
minmax <- employees |>
    group_by(company) |>
    summarize(min = min(total), when_min = month[which.min(total)],
              max = max(total), when_max = month[which.max(total)])
```
```{r, echo = FALSE, results = "asis"}
kbl(minmax) |> kable_styling(full_width = FALSE, position = "center")
```

Here we’ll plot the evolution in time of the fraction of part-time workers
over the total workforce of each company.

```{r}
employees |>
    mutate(part_time = 100 * part_time / total) |>
    ggplot(aes(x = month, y = part_time, colour = company)) +
        geom_line(lwd = 0.8) +
        scale_colour_manual(values = my_palette, labels = full_names) +
        labs(x = "Year", y = "Part-time employees [%]", colour = "Company")
```

## COVID-19 Effect
To analyse the influence of the COVID-19 pandemic, we can zoom in on the
period from 2019 to 2023 in the total workforce’s plot. We also add a 
reference line with the previously-found historic minima of the workforce.

```{r}
employees |>
    dplyr::filter(year(month) > 2018) |>
    ggplot(aes(x = month, y = total, colour = company)) +
        geom_line(lwd = 0.8) +
        geom_hline(aes(yintercept = min, colour = company,
                       linetype = "Historic minimum"),
                   minmax, lwd = 0.6) +
        scale_linetype_manual(name = "", values = rep("dashed", 4)) +
        scale_colour_manual(values = my_palette, labels = full_names) +
        labs(x = "Year", y = "Number of employees", colour = "Company")
```

We see indeed that two out of four companies – Delta and United – took a 
deep hit from the pandemic, resulting in a depletion of the workforce 
almost to an all-time low, especially for Delta. American apparently felt
the crisis a bit less, but we can still observe a dip before the start of
2021.

Federal Express is, on the other hand, in complete contrast with the 
other three companies: its workforce even *grew* in the pandemic period.
This is perhaps not that surprising if we consider that FedEx is a giant
corporation that focuses on basically every aspect of transportation, not 
simply passenger air travel like the other three companies.

# New York City Flights
The data we want to analyse here comes from the `nycflights13` package.

```{r}
library(nycflights13)
library(rcartocolor)
my_palette <- carto_pal(10, "Safe")[1:3]
```

First, let’s have a look at the four tibbles in the dataset.

```{r, echo = FALSE, results = "asis"}
kbl(head(airlines), caption = "`airlines` tibble")
kbl(head(airports), caption = "`airports` tibble")
kbl(head(flights), caption = "`flights` tibble")
kbl(head(planes), caption = "`planes` tibble")
```

## Flights per Day
Now we’ll plot the total number of flights departing from each of the 
three NYC airports (JFK, LGA and EWR) as a function of time.

```{r}
flights$date <- as_date(flights$time_hour)

flights |>
    count(date, origin) |>
    ggplot(aes(x = date, y = n, colour = origin)) +
        geom_point(size = 0.8) +
        geom_smooth(aes(fill = origin), lwd = 0.8) +
        scale_colour_manual(values = my_palette) +
        scale_fill_manual(values = my_palette, guide = "none") +
        labs(x = "Date", y = "Number of flights", colour = "Airport") +
        guides(colour = guide_legend(override.aes = list(fill = NA)))
```

Another plot: this time, the average number of flights per day in a week,
keeping working days and weekends separate, as a function of the week
number in the year.

```{r}
flights |>
    mutate(day_type = ifelse(wday(date) %in% c(1, 7),
                             "weekend", "workday")) |>
    # count by day, keeping info on week and day_type
    count(week = week(date), day, day_type) |>
    # regroup by week and day_type to get the averages we want
    group_by(week, day_type) |>
    summarize(avg = mean(n)) |>
    ggplot(aes(x = week, y = avg, colour = day_type)) +
        geom_line(lwd = 0.8) +
        scale_colour_manual(values = my_palette,
                            labels = c("Weekends", "Workdays")) +
        labs(x = "Week", y = "Average flights per day",
             colour = "Day type")
```

Looking at the workdays’ plot, we can notice a big dip around the 20th and 30th
week, one next to the 50th and another in the last. The last dip is clearly
caused by Christmas and New Year’s eve, while the other two line up perfectly
with the US Independence Day on the 4th of July (`r week("2013-07-04")`th week
of the year) and Thanksgiving, which in 2013 was on the 28th of November 
(`r week("2013-11-28")`th week of the year). Indeed:

```{r, echo = FALSE}
by_week <- flights |>
    dplyr::filter(wday(date) %in% 2:6) |>
    count(week = week(date))

ind_day <- week("2013-07-04")
thnxgiv <- week("2013-11-28")

by_week[(ind_day - 2):(ind_day + 2), ] |>
    kbl(caption = "Flights in the workweeks around July 4th",
        table.attr = "style='width:50%;'") |>
    kable_styling(full_width = TRUE, position = "center")
by_week[(thnxgiv - 2):(thnxgiv + 2), ] |>
    kbl(caption = "Flights in the workweeks around Thanksgiving",
        table.attr = "style='width:50%;'") |>
    kable_styling(full_width = TRUE, position = "center")
```

In the case of Thanksgiving in particular, the decrease in the workday
average is coupled to a sharp increase in the flights over the weekend.
Elsewhere, the weekends present some big dips in some not-so-easily
explainable parts of the calendar, but this is perhaps simply a matter of
less data – we’re averaging over two days instead of five – leading to 
larger fluctuations.

## Departure Delays
Now, let’s examine the trend in the departure delays. We’ll calculate the 
minimum, maximum and average delay for each day of the year, keeping the
three airports’ data separate.
```{r}
dep_delays <- flights |>
    # some flights don’t have a recorded departure and/or arrival time
    drop_na(dep_delay) |>
    group_by(date, origin) |>
    summarize(avg = mean(dep_delay),
              min = min(dep_delay),
              max = max(dep_delay))

plot_delays <- function(key) {
    names <- c(
        "Maximum" = "max",
        "Minimum" = "min",
        "Average" = "avg"
    )

    ggplot(dep_delays, aes(x = date, y = .data[[names[key]]],
                           colour = origin)) +
        geom_point(size = 0.8) +
        geom_smooth(aes(fill = origin), lwd = 0.8) +
        scale_colour_manual(values = my_palette) +
        scale_fill_manual(values = my_palette, guide = "none") +
        labs(x = "Date", y = paste(key, "departure delay [min]"),
            colour = "Airport") +
        guides(colour = guide_legend(override.aes = list(fill = NA)))
}
```

Let’s examine the average delay first.

```{r}
plot_delays("Average")
```

It’s difficult to identify any noteworthy trend – perhaps only a slight 
increase of the delays in the summer months, which is reasonable.
Also, there’s not a significant difference between the three airports.

Now, let’s look at the maxima and minima.

```{r}
plot_delays("Maximum")
plot_delays("Minimum")
```

Nothing too significant here too.

## Plane Speed