---
title: R Laboratory – Exercise 5
author: Guglielmo Bordin
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output:
    prettydoc::html_pretty:
        theme: architect
        highlight: github
        math: katex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    dev = "svg", fig.width = 8, fig.height = 5, message = FALSE
)

set.seed(17052023)
```

# Deaths by horse kicks
Ladislaus Josephovich Bortkiewicz was a Russian economist and statistician.
He noted that the Poisson distribution can be very useful in applied
statistics when describing low-frequency events in a large population. In a
well-known example, he showed that the number of deaths by horse kicks in
the Prussian army followed the Poisson distribution.

Consider the following two sets of observations, taken over a fixed large
time interval in two different corps:

```{r, echo = FALSE}
library(tidyverse)
theme_set(theme_minimal(base_size = 14, base_family = "Open Sans"))

tibble::tribble(
    ~"$y$ dead soldiers", ~"0", ~"1", ~"2", ~"3", ~"4", ~"≥ 5",
    "$n_1$ observations",  109,   65,   22,    3,    1,      0,
    "$n_2$ observations",  144,   91,   32,   11,    2,      0
) |>
    kableExtra::kbl() |>
    kableExtra::kable_styling(full_width = FALSE)
```

We’ll assume a uniform prior, and compute the posterior distribution for 
$\lambda$, the death rate over the measurement time. Let’s define a generic
function for a Bayesian analysis of a Poisson process with a Gamma prior.

```{r}
baypois <- function(prior_alpha, prior_lambda) 1
```
