---
title: R Laboratory – Exercise 2
author: Guglielmo Bordin
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output:
    rmdformats::readthedown:
        fig_width: 6.5
        fig_height: 4
        thumbnails: false
        lightbox: true
        gallery: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(dev = "svg")
```

# Discrete random variable
In this exercise, we’ll examine a probability distribution function of a
discrete random variable $k$ given by
$$
    f(k) =
    \begin{cases}
        k / 15 & \text{when } k = 1, 2, 3, 4, 5, \\
        0      & \text{otherwise}.
    \end{cases}
$$


## PDF and CDF
Now we’ll write the R functions for the probability density $f$ and the
cumulative distribution $F$, following the R naming convention. For the
CDF, we have to sum the probability up to a given threshold:
$$
    F(k) = \sum_{n = 1}^{k} \frac{n}{15} = \frac{k(k + 1)}{30}.  
$$

```{r}
# probability density function
dunif5 <- function(x) {
    ifelse(x %in% 1:5, x / 15, 0)
}

# cumulative distribution function
punif5 <- function(x) {
    ifelse(x > 4, 1, ifelse(x < 1, 0, x * (x + 1) / 30))
}
```

Then, we can view them in a bar plot.
```{r}
# import ggplot and choose font and palette
library(ggplot2)
theme_set(theme_light(base_size = 14, base_family = "Lato"))
my_pal <- c("#83accd", "#20415a", "#dec552", "#7b6220")
```

```{r}
x <- 0:6
# plot the pdf
ggplot() +
    geom_col(aes(x = x, y = dunif5(x)), fill = my_pal[1]) +
    scale_x_continuous(breaks = x) +
    labs(x = "x", y = "f(x)")
```

```{r}
# plot the cdf
ggplot() +
    geom_col(aes(x = x, y = punif5(x)), fill = my_pal[1]) +
    scale_x_continuous(breaks = x) +
    labs(x = "x", y = "F(x)")
```

## Expected values
Having the probability density, we can evaluate the first and second
central moments of the distribution, i.e. the mean and variance:
$$
\begin{aligned}
      \mathrm{E}[k] &= \sum_{k = 1}^{5} k f(k) = \frac{11}{3}, \\
    \mathrm{Var}[k] &= \mathrm{E}[(k - \mu)^2]
                     = \sum_{k = 1}^{5} (k - \mu)^2 f(k) = \frac{14}{9}.
\end{aligned}
$$

```{r}
x <- 1:5
fx <- dunif5(x)

cat("Mean =", mu <- sum(x * fx))
cat("Variance =", sum((x - mu)^2 * fx))
```

Let’s compute another expected value, $\mathrm{E}[k(6 - k)]$.
```{r}
cat("E[k(6 - k)] =", sum(x * (6 - x) * fx))
```

## Sampling from the distribution
Now we implement a function that allows to sample random numbers from the
probability distribution $f(k)$, following the R naming convention with the
`r` prefix. We use the inverse transform method, sampling first from a
uniform distribution (with `runif`), and then choosing the minimum $k$ such
that $F(k)$ is greater than the sampled number.
```{r}
# n = number of samples
runif5 <- function(n) {
    x <- 1:5
    sapply(runif(n), function(r) x[which(punif5(x) > r)[1]])
}
```

Using the newly implemented function, we sample $10^5$ random numbers and
plot them in a normalized histogram superimposed to the PDF.
```{r}
sample <- runif5(1e5)
x <- 1:5
fx <- dunif5(x)

ggplot() +
    geom_histogram(
        aes(x = sample, y = after_stat(density), fill = "hist"),
        binwidth = 1
    ) +
    geom_point(aes(x = x, y = fx, colour = "pdf"), size = 3) +
    geom_line(aes(x = x, y = fx, colour = "pdf"), linewidth = 0.8) +
    scale_fill_manual(values = my_pal[1], labels = "Simulated data") +
    scale_colour_manual(values = my_pal[2], labels = "PDF") +
    labs(x = "x", y = "f(x)", colour = NULL, fill = NULL)
```

# Continuous random variable
Now we’ll analyse the continuous triangular distribution in an interval
$[a, b]$, which is
$$
    f(X) =
    \begin{cases}
        \dfrac{2(x - a)}{(b - a)(c - a)} & a \leqslant x < c, \\[2ex]
        \dfrac{2(b - x)}{(b - a)(b - c)} & c \leqslant x
                                             \leqslant b, \\[2ex]
                                       0 &\text{otherwise}, \\
    \end{cases}
$$
where $c \in [a, b]$.

Let’s define the R function for the PDF, giving some sensible defaults
for the three parameters:
```{r}
dtriang <- function(x, a = 0, c = 0.5, b = 1) {
    ifelse(a <= x & x < c,
           2 * (x - a) / ((b - a) * (c - a)),
           ifelse(c <= x & x <= b,
                  2 * (b - x) / ((b - a) * (b - c)),
                  0))
}
```

… and plot it, choosing $a = 1$, $b = 4$ and $c = 2$.
```{r}
x <- seq(0, 5, by = 0.01)
fx <- dtriang(x, 1, 2, 4)

ggplot() +
    geom_line(aes(x = x, y = fx), colour = my_pal[3], linewidth = 0.8) +
    labs(x = "x", y = "f(x)")
```

## Sampling from the distribution
Once again, let’s write an algorithm to generate random numbers from this
PDF. We start from the CDF: in the interval $[a, c)$ we get
$$
    F_1(x) = \frac{2}{(b - a)(c - a)} \int_a^x (y - a) \, dy 
           = \frac{(x - a)^2}{(b - a)(c - a)},
$$
while in $[c, b]$
$$
    F_2(x) = 1 - \frac{2}{(b - a)(b - c)} \int_x^b (b - y) \, dy
           = 1 - \frac{(b - x)^2}{(b - a)(b - c)}
$$
```{r}
ptriang <- function(x, a = 0, c = 0.5, b = 1) {
    ifelse(a <= x & x < c,
           (x - a)^2 / ((b - a) * (c - a)),
           ifelse(c <= x & x < b,
                  1 - (b - x)^2 / ((b - a) * (b - c)),
                  ifelse(x >= b, 1, 0)))
}

ggplot() +
    geom_line(
        aes(x = x, y = ptriang(x, 1, 2, 4)),
        colour = my_pal[3],
        linewidth = 0.8
    ) +
    labs(x = "x", y = "F(x)")
```

Now let’s invert it, defining the quantile function `qtriang`. With some 
manipulations we can find
$$
\begin{aligned}
    F^{-1}_1(p) &= a + \sqrt{p(b - a)(c - a)}, \\
    F^{-1}_2(p) &= b - \sqrt{(1 - p)(b - a)(b - c)}.
\end{aligned}
$$

```{r}
qtriang <- function(p, a = 0, c = 0.5, b = 1) {
    pc <- (c - a) / (b - a)
    ifelse(p < pc,
           a + sqrt(p * (b - a) * (c - a)),
           b - sqrt((1 - p) * (b - a) * (b - c)))
}

rtriang <- function(n, a = 0, c = 0.5, b = 1) {
    qtriang(runif(n), a, c, b)
}
```

Let’s plot some generated numbers in a histogram to check that everything
works.

```{r}
x <- seq(0, 5, by = 0.01)
fx <- dtriang(x, 1, 2, 4)
sample <- rtriang(1e4, 1, 2, 4)

ggplot() +
    geom_histogram(
        aes(x = sample, y = after_stat(density), fill = "hist"),
        bins = 30
    ) +
    geom_line(aes(x = x, y = fx, colour = "pdf"), linewidth = 0.8) +
    scale_fill_manual(values = my_pal[3], labels = "Simulated data") +
    scale_colour_manual(values = my_pal[4], labels = "PDF") +
    labs(x = "x", y = "f(x)", colour = NULL, fill = NULL)
```

# Waiting at the doctor’s
The waiting time, in minutes, at the doctor’s is about 30 minutes, and the 
distribution follows an exponential PDF with rate $\lambda = 1 / 30$.

The exponential PDF is already provided in R, so we can directly call the
function `rexp` to simulate the waiting time for 60 people, and plot the 
relative histogram.

```{r}
sample <- rexp(60, 1 / 30)

ggplot() +
    geom_histogram(
        aes(x = sample),
        boundary = 0,
        fill = my_pal[1],
        binwidth = 10
    ) +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
    labs(x = "Waiting time (min)", y = "Number of people")
```

The probability for a person to wait less than 12 minutes is given by the
CDF at $x = 12$,
```{r}
cat("P(t <= 12) =", pexp(12, 1 / 30))
```
while for the probability to wait more than one hour,
```{r}
cat("P(t > 60) =", 1 - pexp(60, 1 / 30))
```
We can check if the simulated data returns something similar.
```{r}
cat("Simulated P(t <= 12) =", sum(sample <= 12) / length(sample))
cat("Simulated P(t > 60) =", sum(sample > 60) / length(sample))
```

Now, we evaluate the average waiting time from the simulated data.
```{r}
cat("Sample average =", mean(sample))
```
And we notice that it’s well in line with the expected value from the 
distribution, $1 / \lambda = 30$. We can get the same result by integrating
`x * dexp(x, 1 / 30)` with R.
```{r}
integrate(function(x) x * dexp(x, 1 / 30), 0, Inf)
```

# Multiple choice exams
The final exam of a course has the format of a multiple choice written
test, with five alternatives for each question. A student either knows the
answer, or selects it randomly among the five choices.

If we assume that the probability that a student knows the answer is $p =
0.7$, what is the probability that a student knowingly marks a correct
answer as such? We can use Bayes’ Theorem to answer the question.

In our case, the “data” $D$ is “the student gets the correct answer”, while
the “hypothesis” $H$ is “the student *knew* the correct answer”. We have 
then
$$
    \newcommand{\given}{\,\vert\,}
    P(D \given H) = 1, \quad P(D \given \bar{H}) = 0.2, \quad P(H) = 0.7,
$$
and applying Bayes’ Theorem,
$$
    P(H \given D) = \frac{P(D \given H) P(H)}{P(D \given H) P(H) 
        + P(D \given \bar{H}) P(\bar{H})}
$$

```{r}
cat(1 * 0.7 / (1 * 0.7 + 0.2 * 0.3))
```

Let’s perform a simulation to check the answer.
```{r}
n <- 1e5
correct <- sample(1:5, n, replace = TRUE)
student <- ifelse(runif(n) < 0.7, "knew",
                  ifelse(sample(1:5, 1) == correct, "correct", "wrong"))

cat(sum(student == "knew") / (n - sum(student == "wrong")))
```

# Waiting at the station
Starting from 5:00 in the morning, every half an hour a train departs from
the Milano Centrale station directed to Roma Termini. We assume there is
always an available seat on a train leaving from Milano.

Let’s assume that a person arrives at a random time between 10:45 and
11:45. The distribution of the waiting time is uniform:
```{r}
n <- 1e5
wait_time <- (15 + runif(n, 0, 60)) %% 30

ggplot() +
    geom_histogram(
        aes(x = wait_time),
        binwidth = 1,
        boundary = 0,
        fill = my_pal[3]
    ) +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
    labs(x = "Waiting time (min)", y = "Frequency")
```

What is the probability to wait at most 10 minutes? From the distribution,
we expect
$$
    P(t \leqslant 10) = \frac{10}{30} = \frac{1}{3}.
$$
From the data we get:
```{r}
cat("Simulated P(t <= 10) =", sum(wait_time <= 10) / n)
```

And to wait at least 15 minutes?
$$
    P(t \geqslant 15) = \frac{15}{30} = \frac{1}{2}.
$$
```{r}
cat("Simulated P(t >= 15) =", sum(wait_time >= 15) / n)
```

The average time spent waiting should be 15 minutes.
```{r}
cat("Simulated E[t] =", mean(wait_time))
```

# Stock investment
The annual return rate for a certain stock on the market is a normal
variable with a 10% mean and a 12% standard deviation. Mr X decides to buy
200 shares of the stock at a price of 85€ per share. What is the
probability that after a year his net profit from the investment is at
least 800€? (Supposing there are no transaction costs and annual dividends
for the stock.)

We can first simulate some samples.
```{r}
nsamples <- 1e5
nshares <- 200
price <- 85

sample <- nshares * price * rnorm(nsamples, 0.1, 0.12)
cat("Simulated P(x >= 800) =", sum(sample >= 800) / nsamples)
```
This corresponds to the results we get using the CDF `pnorm`.
```{r}
cat(
    "P(x >= 800) =",
    1 - pnorm(800, 0.1 * price * nshares, 0.12 * price * nshares)
)
```