---
title: R Laboratory – Exercise 4
author: Guglielmo Bordin
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output:
    prettydoc::html_pretty:
        theme: cayman
        highlight: github
        math: katex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    dev = "svg", fig.width = 8, fig.height = 5, message = FALSE
)
```

# Insurance claims
The number of claims received by an insurance company during a week follows
a Poisson distribution with unknown mean $\mu$. The claims per week
observed over a ten week period are:
```
5, 8, 4, 6, 11, 6, 6, 5, 6, 4
```

We’ll start by computing the posterior distribution with the assumption of
a uniform prior for $\mu$. Let’s first define a helpful `bayes` function
to calculate everything we need in the analysis.
```{r}
bayes <- function(lhood, prior, max_mu, cred_int = c(0.025, 0.975)) {
    dm <- 0.001
    m <- seq(0, max_mu, by = dm)
    # get denominator (evidence)
    evid <- integrate(\(x) lhood(x) * prior(x), 0, Inf)$value
    # assemble posterior
    post <- lhood(m) * prior(m) / evid
    cdf <- cumsum(post) * dm

    # get posterior mode, median, mean and std
    mode <- m[which.max(post)]
    median <- m[which.max(cdf > 0.5)]
    mean <- sum(m * post) * dm
    std <- sqrt(sum((m - mean)^2 * post) * dm)

    # get credibility interval
    cred <- sapply(cred_int, \(c) m[which.max(cdf > c)])

    # return named list
    list(
        post = post, mode = mode, med = median,
        mean = mean, std = std, cred = cred
    )
}
```
Now, let’s start with the uniform prior.
```{r}
claims <- c(5, 8, 4, 6, 11, 6, 6, 5, 6, 4)
n <- length(claims)

# poissonian likelihood
lhood <- \(m) m^sum(claims) * exp(-n * m)
# g(mu) = 1 for all mu --> improper prior (divergent integral)
max_mu <- 1.5 * max(claims)
post_unif <- bayes(lhood, \(m) 1, max_mu)
```
```{r, echo = FALSE}
library(tidyverse)

tibble(
    post_unif$mode, post_unif$med, post_unif$mean, post_unif$std,
    paste(post_unif$cred, collapse = ", ")
) |>
    set_names("Mode", "Median", "Mean", "St. dev.", "95% cred. int.") |>
    kableExtra::kbl(caption = "Posterior parameters") |>
    kableExtra::kable_styling(full_width = FALSE)
```
Now, we’ll use a Jeffrey’s prior instead, that is

$$
    g(\mu) = \frac{1}{\sqrt{\mu}}.
$$

This is another improper prior, but it’s not a problem since we normalize
numerically the posterior inside the `bayes` function.

To handle the value at 0 correctly, we also set $g(\mu) = 0$ for $\mu = 0$,
to avoid a divergence in the posterior.
```{r}
post_jeff <- bayes(lhood, \(m) ifelse(m > 0, 1 / sqrt(m), 0), max_mu)
```
```{r, echo = FALSE}
tibble(
    post_jeff$mode, post_jeff$med, post_jeff$mean, post_jeff$std,
    paste(post_jeff$cred, collapse = ", ")
) |>
    set_names("Mode", "Median", "Mean", "St. dev.", "95% cred. int.") |>
    kableExtra::kbl(caption = "Posterior parameters") |>
    kableExtra::kable_styling(full_width = FALSE)
```

Let’s plot the two posteriors with their credibility intervals.
```{r}
my_pal <- c("#EE9C39", "#20A47B", "#5A3920")
theme_set(theme_minimal(base_size = 14, base_family = "Open Sans"))

creds <- list(unif = post_unif$cred, jeff = post_jeff$cred)
prior_names <- c(unif = "Uniform", jeff = "Jeffrey’s")
prior_cols <- my_pal[1:2] |> set_names("unif", "jeff")

tibble(
    mu = seq(0, max_mu, by = 0.001),
    unif = post_unif$post,
    jeff = post_jeff$post,
) |>
    # transform to long format for ggplot
    pivot_longer(-1, names_to = "dist", values_to = "f") |>
    # reorder `dist` factors to follow the order in the text
    mutate(dist = fct_relevel(dist, "unif", "jeff")) |>
    ggplot(aes(x = mu, y = f, colour = dist)) +
        geom_line(linewidth = 0.8) +
        # shade 95% credibility intervals
        geom_area(
            aes(fill = dist),
            data = function(x) {
                x |> group_by(dist) |>
                     filter(mu > creds[[unique(dist)]][1] &
                            mu < creds[[unique(dist)]][2])
            },
            position = "identity",
            alpha = 0.5,
            show.legend = FALSE
        ) +
        scale_colour_manual(values = prior_cols, labels = prior_names) +
        scale_fill_manual(values = prior_cols) +
        labs(
            x = "Poisson parameter μ",
            y = "Posterior distribution P(μ | y)",
            colour = "Prior"
        ) +
        coord_cartesian(expand = FALSE, ylim = c(0, 0.55))
```

Finally, let’s approximate the two posteriors with a normal distribution
with the same mean and standard deviation, and compare the credibility
intervals.
```{r}
tibble(
    c("Uniform", "Jeffrey’s"),
    c(post_unif$mean, post_jeff$mean),
    c(post_unif$std, post_jeff$std),
)
```