---
title: R Laboratory – Exercise 6
author: Guglielmo Bordin
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output:
    prettydoc::html_pretty:
        theme: cayman
        highlight: github
        math: katex
---

```{r setup, include = FALSE}
library(tidyverse)
library(kableExtra)
theme_set(theme_minimal(base_size = 14, base_family = "Open Sans"))
my_pal <- RColorBrewer::brewer.pal(3, "Dark2")

# knitr setup
knitr::opts_chunk$set(
    dev = "svg", fig.width = 8, fig.height = 5,
    message = FALSE, warning = FALSE
)

set.seed(02062023)
```

# COVID-19 and vaccination charts
We’ll analyse the global vaccination dataset provided by “Our World in
Data”, available at ```https://ourworldindata.org/covid-vaccinations```.

```{r}
data <- read_csv("owid-covid-data.csv")
```

The CSV file follows a format of one row per location and date, keeping
track of many variables:

```{r}
spec(data)
```

The first thing that jumps out from a quick analysis of the `iso_code` and
`location` columns is that, among all the countries and territories, there
are also some “grouping” rows, e.g. low income or high income countries, or
continents.

```{r}
data |>
    select(iso_code, location) |>
    filter(str_starts(iso_code, "OWID")) |>
    distinct()
```

So, we could directly work with the `OWID_WRL` rows, and get the
whole-world data. We’ll later make sure that it checks out if we take the
longer route of putting together all the rows pertaining to real single
locations.

We’ll start with the vaccination data: let’s make a quick plot of the
cumulative number of people who received at least one vaccine dose, per
hundred (we find this in the `people_vaccinated_per_hundred` column).

```{r}
data |>
    filter(location == "World") |>
    select(date, people_vaccinated_per_hundred) |>
    drop_na() |>
    ggplot(aes(date, people_vaccinated_per_hundred)) +
        geom_area(
            linewidth = 0.8, alpha = 0.5,
            colour = my_pal[1], fill = my_pal[1]
        ) +
        xlab("Date") +
        ylab("Vaccinated population (%)")
```

Let’s see if we get the same plot by grouping the locations.

```{r}
data_no_owid <- data |> filter(!str_starts(iso_code, "OWID"))

data_no_owid |>
    drop_na(people_vaccinated, population) |>
    group_by(date) |>
    summarize(pvac = 100 * sum(people_vaccinated) / sum(population)) |>
    ggplot(aes(date, pvac)) +
        geom_point(size = 0.8, alpha = 0.5) +
        geom_smooth(
            span = 0.25, linewidth = 0.8,
            colour = my_pal[2], fill = my_pal[2], alpha = 0.5
        ) +
        xlab("Date") +
        ylab("Vaccinated population (%)")
```

As we can see, the trend is indeed similar. We can’t say much about the
discrepancies, not knowing how the `OWID_WRL` data is constructed.

Let’s examine now the distribution of new doses administered by day,
smoothed over a seven-day period. We can find this data in the column
`new_vaccinations_smoothed`.

```{r}
data_no_owid |>
    drop_na(new_vaccinations_smoothed, population) |>
    group_by(date) |>
    summarize(
        nvac = 100 * sum(new_vaccinations_smoothed) / sum(population)
    ) |>
    ggplot(aes(date, nvac)) +
        geom_area(
            linewidth = 0.8, alpha = 0.5,
            fill = my_pal[1], colour = my_pal[1]
        ) +
        xlab("Date") +
        ylab("New vaccinations (% of total population)")
```

We can also try a more interesting analysis by looking at the
“Low/Middle/High Income” rows in the dataset, to see how higher income
countries got hold of the vaccines sooner than the lower income ones. To
avoid the double peaks of the second doses, we’ll look at the variable
`new_people_vaccinated_smoothed`, which filters out the administrations of
second or successive doses.

```{r, fig.width = 10, fig.height = 6}
data |>
    filter(str_ends(location, "income")) |>
    drop_na(new_people_vaccinated_smoothed) |>
    select(date, location, vac = new_people_vaccinated_smoothed) |>
    mutate(location = fct_relevel(
        location,
        "High income",
        "Upper middle income",
        "Lower middle income",
        "Low income"
    )) |>
    ggplot(aes(date, vac)) +
        geom_area(
            aes(date, total),
            data = function(x) {
                x |> group_by(date) |>
                     summarize(total = sum(vac))
            }, colour = "grey40", fill = "grey80", alpha = 0.5
        ) +
        geom_area(
            aes(colour = location, fill = location),
            alpha = 0.5, position = "identity"
        ) +
        scale_y_continuous(
            labels = scales::label_number(suffix = "M", scale = 1e-6)
        ) +
        scale_colour_viridis_d(option = "plasma", guide = "none") +
        scale_fill_viridis_d(option = "plasma") +
        labs(
            x = "Date",
            y = "First doses (7-day smoothed)",
            fill = "Country income"
        )
```

Let’s look at the death count now. First, the cumulative total.

```{r}
data_no_owid |>
    drop_na(total_deaths) |>
    group_by(date) |>
    summarize(deaths = sum(total_deaths)) |>
    ggplot(aes(date, deaths)) +
        geom_area(
            linewidth = 0.8, alpha = 0.5,
            colour = my_pal[3], fill = my_pal[3]
        ) +
        scale_y_continuous(
            breaks = scales::pretty_breaks(),
            labels = scales::label_number(suffix = "M", scale = 1e-6),
            limits = c(0, 8e6)
        ) +
        labs(x = "Date", y = "Total confirmed deaths")
```

And then, the weekly average (from the column `new_deaths_smoothed`).

```{r}
data_no_owid |>
    drop_na(new_deaths_smoothed) |>
    group_by(date) |>
    summarize(deaths = sum(new_deaths_smoothed)) |>
    ggplot(aes(date, deaths)) +
        geom_area(
            linewidth = 0.8, alpha = 0.5,
            colour = my_pal[3], fill = my_pal[3]
        ) +
        scale_y_continuous(
            breaks = scales::pretty_breaks(),
            labels = scales::label_number(suffix = "k", scale = 1e-3)
        ) +
        labs(x = "Date", y = "Daily confirmed deaths (7-day smoothed)")
```